Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]

### 2017-04-24 manuscript/pref_sampling_revision.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

\item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.  However, this assumption is likely violated in many cases when survey effort is non-randomized, leading to preferential sampling.

@@

\item Using simulation, we show that even moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density and that our modeling approach can considerably reduce this bias.

\item We apply our approach to aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the eastern Bering Sea. Models that included a preferential sampling effect led to lower estimates of abundance than models without, but the effect size of the preferential sampling parameter decreased in models that included explanatory environmental covariates.

\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling. Joint models, such as those described in this paper, can be used to test and correct for such biases.  Predictive covariates are also useful for bias reduction, but ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

@@

One of the main advantages of using SDMs is that one is less beholden to predetermined sampling frames and can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}. Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.  For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and an important habitat covariate \citep{AlbertEtAl2010}.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  [Z_i] & = & f(g^{-1}(\mu_i)).
\end{eqnarray}
In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, or spatially autocorrelated random effects.  We use the bracket notation (e.g. $[Z_i]$) to denote a probability mass or probability density function (in the case of discrete or continuous distribution, respectively). In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  [Y_i|Z_i] & = & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_binomial}
  [Y_i|Z_i] & = & \text{Binomial}(Z_i, A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use ``$|$" to specify that a distribution is conditional; for instance, if $[{\bf Y}]$ denotes the marginal probability mass function for ${\bf Y}$, then  $[{\bf Y}|{\bf Z}]$ represents the conditional distribution of ${\bf Y}$ given ${\bf Z}$.  We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.  As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.  We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.  Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using presence only, citizen science, or opportunistically collected data for estimation.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS (e.g. fig. \ref{fig:PS_example}).  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally) , bias may still occur.  Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

In design-based sampling, unequal sampling intensity is often accommodated via stratification or unequal probability sampling, as with Horvitz-Thompson-like estimators where the probability of inclusion varies by sampling unit \citep{Cochran1977}.  However, in the case of PS, this inclusion probability also depends on the value of the response associated with the sampling unit.  Evidently, any approach to account for PS should also account for the dependence between the state variable of interest and the locations chosen for sampling.

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.  For instance, writing observations of the spatial random field at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon_i,
   \label{eq:Z_cont}
\end{eqnarray}
the log-scale relative intensity of sampling locations at $i$ ($\nu_i$) could be written as
\begin{eqnarray}
   \nu_i & \propto & \xi_i + b \mu_i,
   \label{eq:p_cont}
\end{eqnarray}
where $\mu_i$ and $\xi_i$ includes explanatory covariates.
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g. animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates" (see e.g. fig. \ref{fig:PS_example}).    \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn. (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.

@@

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling PS.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the PS model to be written as a simple linear function of the the spatial process model for density.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, sampling may be more strongly preferential close to bases of operations, (e.g. landing strips in the case of aerial surveys), but less so in areas that are harder to get to.

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling. For the abundance process model, we start with eqn. (\ref{eq:process}) as a general formulation for non-Gaussian data, writing the link-scale expectation as in eqn. (\ref{eq:mu}).
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  [R_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from eqn. (\ref{eq:mu}) and ${\bf x}_i^*$ from eqn. (\ref{eq:nu}) may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both eqns. (\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.  In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in eqn. (\ref{eq:nu}) are estimated to permit parameter identification.

 The formulation in eqn. (\ref{eq:nu}) is similar to the one previously proposed for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.  There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf 0}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no PS).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one could contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not previously been investigated.

A comparison of the performance of models with different sets of constraints on ${\bf B}$ can serve as a test of PS.  In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e. not modeling sampling intensity).

@@

For each simulated landscape we generated three virtual count surveys using eqs. \ref{eq:R} and \ref{eq:nu}.  Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.  In the first, we set ${\bf B}=0$, so that surveyed locations were selected independently of the abundance generating process.  For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=3$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.  We refer to these scenarios as moderate and pathological PS, respectively (see fig. \ref{fig:sim_R}).  Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey was set to cover half of the target cell.

We fitted two different models to each count dataset, both of which were provided the habitat covariate used (in part) to generate the data for which a log-linear coefficient $\beta$ was estimated.  In the first model, the elements of \textbf{B} in eqn. (\ref{eq:nu}) were all set to zero.  In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at least when fitted to presence-absence or count data).  In the second model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix. For ease of exposition, we set $p_i = 1.0$ for all simulations (i.e. all animals present were assumed to be detected).

@@

To test for such an effect, we conceptualized bearded seal counts as arising according to the formulation
\begin{eqnarray*}
  [Z_i | \mu_i] & = & \textrm{Poisson}(a_i \exp(\mu_i)) \\
  [Y_i|Z_i] & = & \textrm{Binomial}(Z_i, A_i), \text{ where}
\end{eqnarray*}
$A_i$ defines the proportion of grid cell $i$ that is sampled, $a_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, and $\mu_i$ is defined in eqn. (\ref{eq:mu}).
We modeled the grid cells that were chosen for sampling using eqns. (\ref{eq:R}-\ref{eq:nu}).

We fitted a total of four models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data using the same estimation framework as in the simulation study.  Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) the form of PS ($b=0$ indicates no PS; $b=1$ indicates ${\bf B}=b{\bf I}$, where $b$ is an estimated parameter).  We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but parameter identification was suspect in these models so estimates are not reported here (see Appendix S2 for more information).  When habitat and landscape variables were included, we used three log-linear predictors: linear and quadratic functions of sea-ice concentration, and distance from the southern ice edge.  Remotely sensed sea-ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}.  Models for $\mu_i$ and $\nu_i$ both utilized spatially autocorrelated random effects with a Mat\'{e}rn covariance function between grid cell centroids (see Appendix S1 for further details).  When covariates were included, they were included in both models (i.e. for $\mu_i$ and $\nu_i$).

@@

%http://www.plosone.org/article/info:doi/10.1371/journal.pone.0036527

>>>>

\item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.  However, this assumption may be violated when survey effort is non-randomized, leading to preferential sampling.

@@

\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density and that our modeling approach can considerably reduce this bias.  In contrast, preferential sampling did not appear to bias inferences about parameters informing species-habitat relationships (i.e. slope parameters).

\item We apply our approach to aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the eastern Bering Sea. As expected, models with a preferential sampling effect led to lower abundance than those without. However, several lines of reasoning (better predictive performance, higher biological realism) led us to prefer models without a preferential sampling effect for this data set.

\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling when estimating animal density or occurrence. Joint models, such as those described in this paper, can be used to test and correct for such biases. However, such models can be unstable; ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
\end{enumerate}

@@

One of the main advantages of using SDMs is that one is less beholden to predetermined sampling frames and can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}. Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.  For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and important habitat covariates \citep{AlbertEtAl2010}.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  [Z_i] & = & f(g^{-1}(\mu_i)),
\end{eqnarray}
where we use the bracket notation ($[\cdot]$) to denote a probability distribution (density) function.  In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, and spatially autocorrelated random effects.  Spatially autocorrelated random effects are often included to allow animal density to vary smoothly in space and to account for residual patchiness not explained by predictive covariates \citep{Legendre1993,LichsteinEtAl2002}. In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{ReichEtAl2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to conduct a complete census of all animals present in a surveyed location, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable is $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  [Y_i|Z_i] & = & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates (e.g. visibility, observer experience). Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_binomial}
  [Y_i|Z_i] & = & \text{Binomial}(Z_i, A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use ``$|$" to specify that a distribution is conditional; for instance, if $[{\bf Y}]$ denotes the marginal probability mass function for ${\bf Y}$, then  $[{\bf Y}|{\bf Z}]$ represents the conditional distribution of ${\bf Y}$ given ${\bf Z}$.  We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.  For animal abundance, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.  We use the notation $Z_i$ when describing the process model in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

Design-based estimation requires that surveys adhere to a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.  Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using presence-only, citizen science, or opportunistically collected data for estimation.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS (e.g. fig. \ref{fig:PS_example}).  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally), bias may still occur.  Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

In design-based sampling, unequal sampling intensity is often accommodated via stratification or unequal probability sampling, as with Horvitz-Thompson-like estimators where the probability of inclusion varies by sampling unit \citep{Cochran1977}.  However, in the case of PS, this inclusion probability also depends on the value of the response associated with the sampling unit.  Evidently, any approach to account for PS should also account for the dependence between the process of interest and the locations chosen for sampling.

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models on a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the process of interest and the locations chosen for sampling.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity at a given location depended linearly on the process of interest at that location.  For instance, writing observations of the spatial process (random field) at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon_i,
   \label{eq:Z_cont}
\end{eqnarray}
the log-scale relative intensity of sampling locations at $i$ ($\nu_i$) could be written as
\begin{eqnarray}
   \nu_i & \propto & \xi_i + b \mu_i,
   \label{eq:p_cont}
\end{eqnarray}
where $\mu_i$ and $\xi_i$ can both be written as a function of predictive covariates.
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g. animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates" (see e.g. fig. \ref{fig:PS_example}).    \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, log sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn. (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.

@@

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling PS.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the PS model to be written as a simple linear function of the spatial process.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, sampling may be more strongly preferential close to bases of operations, (e.g. landing strips in the case of aerial surveys), but less so in areas that are harder to get to.

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling. For the abundance process model, we start with eqn. (\ref{eq:process}) as a general formulation for non-Gaussian data, writing the link-scale expectation as in eqn. (\ref{eq:mu}).
Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
\begin{eqnarray}
 \label{eq:R}
  [R_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
\label{eq:nu}
\end{eqnarray}
  As with the process model, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).  The predictive covariates ${\bf x}_i$ from eqn. (\ref{eq:mu}) and ${\bf x}_i^*$ from eqn. (\ref{eq:nu}) may or may not be the same.  Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both eqns. (\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between sampling intensity and the process of interest.  The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.

 The formulation in eqn. (\ref{eq:nu}) is similar (but not identical) to one proposed by \citet{RoyleBerliner1999} for hierarchical multivariate models with spatial dependence.  There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf 0}_{S \times S}$ (i.e. to an all-zero matrix) corresponds to an absence of spatial dependence (and thus no PS).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one could contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by
 \begin{eqnarray*}
   \label{eq:ts}
   b_{ii} & = & b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i,
 \end{eqnarray*}
 where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  To our knowledge, the identifiability of higher order models (such as trend surfaces) has not been investigated.

A comparison of the performance of models with different sets of constraints on ${\bf B}$ can serve as a test of PS.  In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs.

@@

For each simulated landscape we generated three virtual count surveys using eqs. \ref{eq:R} and \ref{eq:nu}.  Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects on the probability of sampling), but differed in how the matrix ${\bf B}$ was parameterized.  In the first, we set ${\bf B}=0$, so that surveyed locations were randomly selected (i.e. independent of abundance).  For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=3$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the spatial abundance process in that unit.  We refer to these scenarios as moderate and pathological PS, respectively (see fig. \ref{fig:sim_R}).  Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey covered half of each target cell.

We fitted two different models to each count dataset. In the first model, the elements of \textbf{B} in eqn. (\ref{eq:nu}) were all set to zero.  In this case, sampling intensity is assumed to be independent of underlying abundance, as implicitly assumed by most SDMs in the ecological literature.  In the second model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix. For ease of exposition, we set $p_i = 1.0$ for all simulations (i.e. all animals present were assumed to be detected).  Each model included the same habitat covariate that was used to generate the data as a log-linear predictor of abundance.

@@

To test for such an effect, we conceptualized bearded seal counts as arising according to the formulation
\begin{eqnarray*}
  [Z_i | \mu_i] & = & \textrm{Poisson}(a_i \exp(\mu_i)) \\
{}  [Y_i|Z_i] & = & \textrm{Binomial}(Z_i, A_i), \text{ where}
\end{eqnarray*}
$A_i$ defines the proportion of grid cell $i$ that is sampled, $a_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, and $\mu_i$ is defined in eqn. (\ref{eq:mu}).
We modeled the grid cells that were chosen for sampling using eqns. (\ref{eq:R}-\ref{eq:nu}).

We fitted a total of six models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1},M_{cov=1,b=ts},M_{cov=1,b=ts}$) to bearded seal count data using the same estimation framework as in the simulation study.  Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) the form of PS ($b=0$ indicates no PS; $b=1$ indicates ${\bf B}=b{\bf I}$; $b=ts$ indicates a trend surface specification for PS, see eqn. \ref{eq:ts}).  When habitat and landscape variables were included, we used the following log-linear predictors for the abundance process: linear and quadratic functions of sea-ice concentration, linear and quadratic functions of depth, distance from land, and distance from the southern ice edge. In contrast, distance from the Alaska mainland was the only predictor used for sampling intensity ($\bfnu$). Remotely sensed sea-ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}. All measurements were made relative to grid cell centroids.  Models for $\mu_i$ and $\nu_i$ both utilized spatially autocorrelated random effects with a Mat\'{e}rn covariance function (see Appendix S1 for further details).

To compare and contrast models, we calculated marginal AIC and computed mean square prediction error. Prediction error was calculated using a Monte Carlo cross validation procedure.  For a total of 100 replicates, we randomly selected 40/394 counts to be withheld for testing.  Estimation proceeded using the remaining 354 counts for each of the six estimation models; mean squared error of observed and predicted counts was used as the scoring function for calculation of prediction error with each test data set \citep{HootenHobbs2015}.  Note that in each cross validation analysis that we used $R_i$ values from the full data set (i.e. count predictions were made assuming these cells had been sampled).

@@
